---
title: "Transparency of COVID-19-related research in dental journals - Database creation"
author: Ahmad Sofi-Mahmudi, Eero Raittio
date: "2022-01-26"
output: html_document
---

# Loading required packages

```{r}
pacman::p_load(tidyverse,
               rtransparent, 
               metareadr, 
               europepmc)
```


# Load the datasets

## Load International Standard Serial Numbers (ISSNs)

```{r}
ISSNs <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTcoDSEDhtzrYe8u8hkLC9OQ9fZHaQn8cBNZt_7k-qnqtDFP0OCxXPj9aM38D0umO2stdox41C1eZAi/pub?gid=0&single=true&output=csv")
```

## Some modifications are needed...
```{r}
ISSNs <- ISSNs %>% mutate(search.term = paste0("ISSN:", Primary.ISSN))
ISSNsQuery <- paste(ISSNs$search.term, 
                    collapse = " OR ")
```

## Load all open access dentistry and oral health papers indexed in PubMed to the end of 2021:

```{r}
COVID_19_keywords <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vS6ibaTIJ0-4MiOZrLHKwo-Ya3MjlGX4f-dPQwdmWwQZhRGi6C83K1I_WceUJPH9zGyBc4L-lSalET_/pub?output=csv")
```

```{r}
COVID_19_keywords <- COVID_19_keywords %>% mutate(
  search.term.title = paste0("TITLE:", '"', Keyword, '"'),
  search.term.keyword = paste0("KW:", '"', Keyword, '"'),
  search.term.results = paste0("RESULTS:", '"', Keyword, '"'))

TitleQuery <- paste(COVID_19_keywords$search.term.title,
                     collapse = " OR ")

KeywordQuery <- paste(COVID_19_keywords$search.term.keyword,
                     collapse = " OR ")

ResultQuery <- paste(COVID_19_keywords$search.term.results,
                     collapse = " OR ")
```


```{r}
db <- epmc_search(
  query = paste0(
    "'(",
    ISSNsQuery,
    ") AND",
    " (",
    TitleQuery, " OR ", KeywordQuery, " OR ", ResultQuery,
    ") ",
    'AND (SRC:"MED") 
    AND (LANG:"eng" OR LANG:"en" OR LANG:"us") 
    AND (FIRST_PDATE:[2020-01-01 TO 2021-12-31])
    AND ((IN_EPMC:y) OR (OPEN_ACCESS:y))
    AND (PUB_TYPE:"Journal Article" OR PUB_TYPE:"research-article" OR PUB_TYPE:"rapid-communication" OR PUB_TYPE:"product-review")',
    "'"
  ),
  limit = 100000
)
```


Let's see the first six rows of the database:
```{r}
head(db)
```

Removing "PMC" from the cells
```{r}
db$pmcid_ <- gsub("PMC", "", as.character(db$pmcid))
```

Now, we make a folder for xml format articles and switch to that folder:
```{r}
dir.create("pmc")

setwd("pmc")
```

Next, we download xmls in format accessible with rtransparent:
```{r}
db <- subset(db, !(pmcid_ %in% c(8028467, 8013335, 8033587, 7780260, 8463922, 7342805, 7530803))) #removed due to the error "The metadata format 'pmc' is not supported by the item or by the repository."

sapply(db$pmcid_, mt_read_pmcoa)
```
Error in 8013644 - The metadata format 'pmc' is not supported by the item or by the repository.

Error in 7572235 - The value of the identifier argument is unknown or illegal in this repository.

Now we run rtransparent:
```{r}
filepath = dir(pattern=glob2rx("PMC*.xml"))

results_table_all <- sapply(filepath, rt_all_pmc)

results_table_data <- rt_data_code_pmc_list(
  filepath,
  remove_ns=F,
  specificity = "low")
```

A list is created now. We should convert this list to a dataframe:
```{r}
df <- data.table::rbindlist(results_table_all, fill = TRUE)
```


Merge data sharing results to database file:

```{r}
setwd('..')
opendata <- merge(db, results_table_data, by = "pmid") %>% merge(df)
View(opendata)
```


Random sample of search results for validation of our approach (100 refs):
```{r}
set.seed(100)

randomsample <- db[sample(nrow(db), 100), ]
```

