---
title: "Transparency of COVID-19-related research in dental journals - Analysis"
author: Ahmad Sofi-Mahmudi, Eero Raittio
date: "2022-01-26"
output: html_document
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/ahmad/Google Drive/Meta-Research in Dentistry/Data Transparency Project/COVID-19 dental data/data/')
```

# Loading required packages

```{r}
pacman::p_load(tidyverse, 
               europepmc)
```

# Loading the datasets
First, set working directory to the folder that has all the datasets in it using setwd() function and then run the following code:
```{r}
opendata <- read.csv("covid_dental_transparent_opendata.csv")
```

Then, we will read the file that containts ISSNs:
```{r}
ISSNs <- read.csv("dental_journals_issns.csv")
```

Some modifications are needed to be done to the ISSNs dataframe:
```{r}
ISSNs <- ISSNs %>% mutate(search.term = paste0("ISSN:", Primary.ISSN))
ISSNsQuery <- paste(ISSNs$search.term, 
                    collapse = " OR ")
```

Now, we need COVID-19 keywords. Loading its csv file:
```{r}
COVID_19_keywords <- read.csv("covid_keywords.csv")
```

Some modifications is needed to reach the queries:
```{r}
COVID_19_keywords <- COVID_19_keywords %>% mutate(
  search.term.title = paste0("TITLE:", '"', Keyword, '"'),
  search.term.keyword = paste0("KW:", '"', Keyword, '"'),
  search.term.results = paste0("RESULTS:", '"', Keyword, '"'))

TitleQuery <- paste(COVID_19_keywords$search.term.title,
                     collapse = " OR ")

KeywordQuery <- paste(COVID_19_keywords$search.term.keyword,
                     collapse = " OR ")

ResultQuery <- paste(COVID_19_keywords$search.term.results,
                     collapse = " OR ")
```

# Results
Number of all papers (open access and non open access):
```{r}
hits_all <- epmc_hits(
  query = paste0(
    "'(",
    ISSNsQuery,
    ") AND",
    " (",
    TitleQuery, " OR ", KeywordQuery, " OR ", ResultQuery,
    ") ",
    'AND (SRC:"MED") 
    AND (LANG:"eng" OR LANG:"en" OR LANG:"us") 
    AND (FIRST_PDATE:[2020-01-01 TO 2021-12-31])
    AND (PUB_TYPE:"Journal Article" OR PUB_TYPE:"research-article" OR PUB_TYPE:"rapid-communication" OR PUB_TYPE:"product-review")',
    "'"
  ))

hits_all
```

Number of open access papers:
```{r}
hits_oa <- epmc_hits(
  query = paste0(
    "'(",
    ISSNsQuery,
    ") AND",
    " (",
    TitleQuery, " OR ", KeywordQuery, " OR ", ResultQuery,
    ") ",
    'AND (SRC:"MED") 
    AND (LANG:"eng" OR LANG:"en" OR LANG:"us") 
    AND (FIRST_PDATE:[2020-01-01 TO 2021-12-31])
    AND ((IN_EPMC:y) OR (OPEN_ACCESS:y))
    AND (PUB_TYPE:"Journal Article" OR PUB_TYPE:"research-article" OR PUB_TYPE:"rapid-communication" OR PUB_TYPE:"product-review")',
    "'"
  ))

hits_oa
```

Open access percentage:
```{r}
round((hits_oa/hits_all)*100, 2)
```

Number of open acess papers per year:
```{r}
table(opendata$year_epub)
```

The number of journals in our dataset:
```{r}
length(table(opendata$journal))
```
Journals with the highest number of articles in our dataset, from high to low:
```{r}
table(opendata$journal) %>% as.data.frame() %>% arrange(desc(Freq))
```
The mean and the median of the number of citations to these references:
```{r}
data.frame(Mean = round(mean(opendata$citedByCount),1),
           SD = round(sd(opendata$citedByCount), 2),
           Median = median(opendata$citedByCount),
           IQR = IQR(opendata$citedByCount)
           )

```
Characteristics of the paper with the highest number of citations:
```{r}
opendata[which.max(opendata$citedByCount),] %>% 
  select(pmid, pmcid, doi.x, title, authorString, journalTitle, pubYear)
```

The number and percentage of papers with a conflict of interest (CoI) disclosure:
```{r}
data.frame(
  number = length(opendata$is_coi_pred[opendata$is_coi_pred == TRUE]),
           percentage = round(length(opendata$is_coi_pred[opendata$is_coi_pred == TRUE])/nrow(opendata)*100, 2)
           )
```

Confidence interval for CoI:
```{r}
library(epiR)

round(epi.prev(pos = length(opendata$is_coi_pred[opendata$is_coi_pred == TRUE]),
         tested = nrow(opendata),
         se = 0.992,
         sp = 0.995)$ap, 
      2)
```

Comparing CoI disclosure rate in two years:
```{r}
round(prop.table(table(opendata$is_coi_pred, opendata$year_epub), 2)*100, 1)
```

Chi-square test:
```{r}
chisq.test(opendata$is_coi_pred, opendata$year_epub)
```

The number and percentage of papers with a funding statement:
```{r}
data.frame(number = length(opendata$is_fund_pred[opendata$is_fund_pred == TRUE]),
           percentage = round(length(opendata$is_fund_pred[opendata$is_fund_pred == TRUE])/nrow(opendata)*100, 2)
           )
```

Confidence interval for funding statement:
```{r}

round(epi.prev(pos = length(opendata$is_fund_pred[opendata$is_fund_pred == TRUE]),
         tested = nrow(opendata),
         se = 0.997,
         sp = 0.981)$ap, 
      2)
```

Comparing funding statement rate in two years:
```{r}
round(prop.table(table(opendata$is_fund_pred, opendata$year_epub), 2)*100, 1)
```

Chi-square test:
```{r}
chisq.test(opendata$is_fund_pred, opendata$year_epub)
```

The number and percentage of papers that were registered beforehand:
```{r}
data.frame(number = length(opendata$is_register_pred[opendata$is_register_pred == TRUE]),
           percentage = round(length(opendata$is_register_pred[opendata$is_register_pred == TRUE])/nrow(opendata)*100, 2)
           )
```

Confidence interval for registration:
```{r}

round(epi.prev(pos = length(opendata$is_register_pred[opendata$is_register_pred == TRUE]),
         tested = nrow(opendata),
         se = 0.955,
         sp = 0.997)$ap, 
      2)
```

Comparing registration rate in two years:
```{r}
round(prop.table(table(opendata$is_register_pred, opendata$year_epub), 2)*100, 1)
```

Chi-square test:
```{r}
chisq.test(opendata$is_register_pred, opendata$year_epub)
```

The number and percentage of papers that shared data (preliminary):
```{r}
data.frame(number = length(opendata$is_open_data[opendata$is_open_data == TRUE]),
           percentage = round(length(opendata$is_open_data[opendata$is_open_data == TRUE])/nrow(opendata)*100, 2)
           )
```

Comparing data sharing rate in two years (preliminary):
```{r}
round(prop.table(table(opendata$is_open_data, opendata$year_epub), 2)*100, 1)
```

The number and percentage of papers that shared code (preliminary):
```{r}
data.frame(number = length(opendata$is_open_code[opendata$is_open_code == TRUE]),
           percentage = round(length(opendata$is_open_code[opendata$is_open_code == TRUE])/nrow(opendata)*100, 2)
           )
```

Comparing code sharing rate in two years (preliminary):
```{r}
round(prop.table(table(opendata$is_open_code, opendata$year_epub), 2)*100, 1)
```

Adding journal Impact Factors to the dataset:
```{r}
opendata <- merge(x = opendata, 
                  y = ISSNs[, c("Abbreviation", "X2020.JIF")],
                  by.x = "journalTitle",
                  by.y = "Abbreviation",
                  all.x = TRUE)
```

Total number of papers published in journals with no JIF:
```{r}
sum(is.na(opendata$X2020.JIF))
```

```{r}
opendata %>% 
  group_by(is_coi_pred) %>% 
  summarise(Mean = round(mean(X2020.JIF, na.rm = T),3),
            SD = round(sd(X2020.JIF, na.rm = T),4),
            Median = median(X2020.JIF, na.rm = T),
            IQR = IQR(X2020.JIF, na.rm = T))
```

Test for normality:
```{r}
with(opendata, shapiro.test(X2020.JIF[is_coi_pred == TRUE]))
with(opendata, shapiro.test(X2020.JIF[is_coi_pred == FALSE]))
```

Data is not normal, hence we will use unpaired two-samples Wilcoxon test.
```{r}
wilcox.test(X2020.JIF~is_coi_pred, data = opendata, exact = FALSE)
```

```{r}
opendata %>% 
  group_by(is_fund_pred) %>% 
  summarise(Mean = round(mean(X2020.JIF, na.rm = T), 3),
            SD = round(sd(X2020.JIF, na.rm = T), 4),
            Median = median(X2020.JIF, na.rm = T),
            IQR = IQR(X2020.JIF, na.rm = T))
```

Test for normality:
```{r}
with(opendata, shapiro.test(X2020.JIF[is_fund_pred == TRUE]))
with(opendata, shapiro.test(X2020.JIF[is_fund_pred == FALSE]))
```

Data is not normal, hence we will use unpaired two-samples Wilcoxon test.
```{r}
wilcox.test(X2020.JIF~is_fund_pred, data = opendata, exact = FALSE)
```

```{r}
opendata %>% 
  group_by(is_register_pred) %>% 
  summarise(Mean = round(mean(X2020.JIF, na.rm = T), 3),
            SD = round(sd(X2020.JIF, na.rm = T), 4),
            Median = median(X2020.JIF, na.rm = T),
            IQR = IQR(X2020.JIF, na.rm = T))
```

Test for normality:
```{r}
with(opendata, shapiro.test(X2020.JIF[is_register_pred == TRUE]))
with(opendata, shapiro.test(X2020.JIF[is_register_pred == FALSE]))
```

Data is not normal, hence we will use unpaired two-samples Wilcoxon test.
```{r}
wilcox.test(X2020.JIF~is_register_pred, data = opendata, exact = FALSE)
```


## Monthly trend
Adding month column to dataset:
```{r}
library(lubridate)

opendata <- opendata %>% 
  mutate(month = month(
    as.POSIXlt(firstPublicationDate, 
               format = "%m/%d/%Y")),
    year = year(
      as.POSIXlt(firstPublicationDate, 
               format = "%m/%d/%Y")
    ))
```

Now, drawing monthly trend for each year:
```{r}
summarybyMonth <- opendata %>%
  group_by(month, year) %>%
  summarise(counts = n(),
            coi_true = sum(is_coi_pred == TRUE),
            fund_true = sum(is_fund_pred == TRUE),
            reg_true = sum(is_register_pred == TRUE))

summarybyMonth %>%
  ggplot(aes(x = month, y = counts)) + 
  geom_col(position = position_dodge()) +
  facet_grid(year ~ .) +
  geom_bar(stat = "identity") +
  scale_x_discrete(name="Month", limits = 1:12, labels = 1:12)


coi_summarybyMonth <- opendata %>%
  group_by(month, year, is_coi_pred) %>%
  summarise(counts = n())

coi_plot <- coi_summarybyMonth %>%
  ggplot(aes(x = month, y = counts, fill = is_coi_pred)) + 
  geom_col(position = position_dodge()) +
  facet_grid(. ~ year) +
  geom_bar(stat = "identity") +
  scale_x_discrete(name="Month", limits = 1:12, labels = 1:12) +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  labs(title = "Conflict of Interest disclosure")

fund_summarybyMonth <- opendata %>%
  group_by(month, year, is_fund_pred) %>%
  summarise(counts = n())

fund_plot <- fund_summarybyMonth %>%
  ggplot(aes(x = month, y = counts, fill = is_fund_pred)) + 
  geom_col(position = position_dodge()) +
  facet_grid(. ~ year) +
  geom_bar(stat = "identity") +
  scale_x_discrete(name="Month", limits = 1:12, labels = 1:12) +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  labs(title = "Funding statement")

reg_summarybyMonth <- opendata %>%
  group_by(month, year, is_register_pred) %>%
  summarise(counts = n())

reg_plot <- reg_summarybyMonth %>%
  ggplot(aes(x = month, y = counts, fill = is_register_pred)) + 
  geom_col(position = position_dodge()) +
  facet_grid(. ~ year) +
  geom_bar(stat = "identity") +
  scale_x_discrete(name="Month", limits = 1:12, labels = 1:12) +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  labs(title = "Registration statement")

library(ggpubr)
figure <- ggarrange(coi_plot + rremove("xlab") + theme(axis.text.x=element_blank()), 
                    fund_plot + rremove("xlab") + theme(axis.text.x=element_blank(), strip.text.x = element_blank()),
                    reg_plot + theme(strip.text.x = element_blank()),
                    ncol = 1, nrow = 3, vjust = 1, 
                    align = "hv", common.legend = T, legend = "right")

# tiff("Figure.tiff", width = 6, height = 7, units = "in", res = 300)
figure
#dev.off()
```


